---
title: "Using heatmaply with gene expression data - Visualization of raw data (median-centered data, PAM50 genes only)"
author: "Alan O'Callaghan"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
    toc: true
    fig_width: 12
    fig_height: 10
    depth: 3  # upto three depths of headings (specified by #, ## and ###)  
    number_sections: true  ## if you want number sections at each table header
    theme: yeti  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Using heatmaply with gene expression data - Visualization of raw data (median-centered data, PAM50 genes only)}
-->


```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
   # cache = TRUE,
   dpi = 60,
  comment = '#>',
  tidy = FALSE)

```
**Author**: Alan O'Callaghan (alan.b.ocallaghan@gmail.com)




Introduction
============

Due to the size of the objects, this file is seperated to three fils. You can view this series in the following links:

* [Using heatmaply with gene expression data - Visualization of raw and voom-transformed data (all genes)](https://cdn.rawgit.com/talgalili/heatmaplyExamples/master/inst/doc/biological_data.html)
* [Using heatmaply with gene expression data - Visualization of raw data (median-centered data, PAM50 genes only)](https://cdn.rawgit.com/talgalili/heatmaplyExamples/master/inst/doc/biological_data_2.html)
* [Using heatmaply with gene expression data - Visualization of voom-transformed data (median-centered data, PAM50 genes only)](https://cdn.rawgit.com/talgalili/heatmaplyExamples/master/inst/doc/biological_data_3.html)

This is file 2 in the series.


```{r}
# Let's load the packages
library(heatmaply)
library(heatmaplyExamples)
```



Visualization of raw data (median-centered data, PAM50 genes only)
=================================================================


Breast cancer has been studied extensively using gene expression profiling methods (see [Breast cancer gene expression]). This has lead to the identification of a number of gene sets which stratify patients into molecular subgroups.
One such gene set is commonly known as the [PAM50] gene set. In this example, we will visualize gene expression patterns using heatmaply.

Centering data before performing clustering tends to result in more meaningful cluster
assignment. This is particularly true when the measure of interest is the similarity in patterns across features, rather than the total distance between values. Furthermore,
it is typically the difference between samples which is of interest, rather than the difference between measures. Non-centered data may show that all samples measure high for one variable, and low for another, while centered data shows relative differences. Alternatively, one could use a distance measure which is invariant to total distance, such as correlation. 
Heatmaps of non-centered are shown in 
[another vignette](non_centred_heatmaps.html) within this package. It can be seen
that the concordance between the centered and non-centered heatmaps is mediocre, and 
the clustering of triple negative samples is not as definite.

In the heatmaps shown below, it is clear that samples appear to cluster loosely 
based on [PAM50] subtype more than the previous examples.
Concordance with the assigned labels shown in the row annotation 
is not complete, however this may be expected, 
given that a different clustering method was used here (hierarchical clustering,
rather than [k-medioids]).


```{r}
pam50_genes <- intersect(pam50_genes, rownames(raw_expression))
raw_pam50_expression <- raw_expression[pam50_genes, ]
voomed_pam50_expression <- voomed_expression[pam50_genes, ]

center_raw_mat <- cor_mat_raw_logged - 
    apply(cor_mat_raw_logged, 1, median)

raw_max <- max(abs(center_raw_mat), na.rm=TRUE)
raw_limits <- c(-raw_max, raw_max)

```


```{r}

heatmaply(t(center_raw_mat), 
    row_side_colors = tcga_brca_clinical,
    showticklabels = c(FALSE, FALSE),
    fontsize_col = 7.5,
    col = cool_warm(100),
    main = 'Centred log2 read counts, PAM50 genes',
    limits = raw_limits,
    plot_method = 'plotly')


```

Note that `heatmaply_cor` is just like `heatmaply` but with defaults that are better suited for correlation matrix (limits from -1 to 1, and a cold-warm color scheme).

```{r, fig.width=13, fig.height=10}
heatmaply_cor(cor(center_raw_mat), 
    row_side_colors = tcga_brca_clinical,
    showticklabels = c(FALSE, FALSE),
    main = 'Sample-sample correlation based on centred, log2 PAM50 read counts',
    plot_method = 'plotly')

```






Discussion
==========
It may be useful when examining expression 
heatmaps to identify particularly high or low measures for a single
gene in a group of patients, or a gene which shows unusually high or low variance.
The mouse-over text available in the `heatmaply` 
package allows visual assessment of measures of interest and quick identification
of samples or genes with unusual gene expression patterns.
Similarly, visualizing correlation heatmaps with `heatmaply` allows the user to
rapidly identify samples with unusually high or low pairwise correlation.

References
==========
- [limma]
- [voom](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29)
- [Trimmed mean of M-component normalisation][TMM]
- [Genomics Data Commons]
- [PAM50]
- [Breast cancer gene expression]
- [Partitioning around medioids]

[Partitioning around medioids]: https://en.wikipedia.org/wiki/K-medoids
[k-medioids]: https://en.wikipedia.org/wiki/K-medoids
[Genomics Data Commons]: https://gdc.cancer.gov/
[limma]: https://bioconductor.org/packages/release/bioc/html/limma.html
[TMM]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2864565/
[PAM50]: http://ascopubs.org/doi/abs/10.1200/jco.2008.18.1370
[Genomic Data Commons]: https://gdc.cancer.gov/
[data preprocessing vignette]: data_preprocessing.html
[voom]: https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29
[Breast cancer gene expression]: https://www.ncbi.nlm.nih.gov/pubmed/28733194




sessionInfo
===============

```{r}
sessionInfo()
```


