---
title: "Using heatmaply with the measles data set"
author: "Tal Galili"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: yes
    toc: true # table of content true
    fig_width: 6
    fig_height: 6
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: yeti  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Using heatmaply with the measles data set}

-->

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
knitr::opts_chunk$set(
   # cache = TRUE,
   dpi = 60,
  comment = '#>',
  tidy = FALSE)


```






```{r}
library(heatmaply)
library(dendextend)
```


Measles
=============

How the data was prepared 
------------------------

Here `eval=FALSE` as we have the final version in the heatmaplyExamples package.


Preparing the data:

```{r, eval=FALSE}

# measles
# bil gates (20 Feb 2015): https://twitter.com/BillGates/status/568749655112228865
# http://graphics.wsj.com/infectious-diseases-and-vaccines/#b02g20t20w15
# http://www.opiniomics.org/recreating-a-famous-visualisation/
# http://www.r-bloggers.com/recreating-the-vaccination-heatmaps-in-r-2/




# source: http://www.r-bloggers.com/recreating-the-vaccination-heatmaps-in-r/
# measles <- read.csv("https://raw.githubusercontent.com/blmoore/blogR/master/data/measles_incidence.csv", header=T, 
#                     skip=2, stringsAsFactors=F)
# dir.create("data")
# write.csv(measles, file = "data\\measles.csv")
# measles[measles == "-"] <- NA
measles[,-(1:2)] <- apply(measles[,-(1:2)], 2, as.numeric)

dim(measles)
# head(measles)

d2 <- aggregate(measles, list(measles[, "YEAR"]), "sum", na.rm = T)
rownames(measles) <- d2[,1]
d2 <- d2[, -c(1:4)]
d2 <- t(measles)
# head(measles)

dim(measles)
d2[1:5,1:5]

# head(md2)
# 
# # per week
# d22 <- aggregate(measles, list(measles$YEAR, measles$WEEK), "sum", na.rm = T)
# rownames(measles2) <- paste(measles2[,1], d22[,2], sep = "_")
# d22 <- d22[, -c(1:5)]
# d22 <- t(measles2)
# md22 <- reshape2::melt(measles2)
# dim(measles2) # 51 3952


```

```{r, eval=FALSE}

# pander::pander(sqrt(measles))
if(!file.exists("data\\sqrt_d2.csv")) write.csv(sqrt(measles), file = "data\\sqrt_d2.csv")
```

Let's have measles in a long format.


```{r}
library(heatmaplyExamples)
data(measles)
dim(measles)
md2 <- reshape2::melt(measles)
head(md2)
```


Which transformation to use?
-------------

We can look at the general timeline trend with different transformations.


```{r}

library(ggplot2)
ggplot(md2, aes(x = Var2, y = value)) + geom_line(aes(group = Var1)) + geom_smooth(size = 2)
ggplot(md2, aes(x = Var2, y = sqrt(value))) + geom_line(aes(group = Var1)) + geom_smooth(size = 2)
ggplot(md2, aes(x = Var2, y = log(value+.1))) + geom_line(aes(group = Var1)) + geom_smooth(size = 2)



# ggplot(md22, aes(x = Var2, y = value)) + geom_line(aes(group = Var1)) # + geom_smooth(size = 2)

#
# we miss the per state information and the patters of similarity



```


We can try several heatmaps settings until we get one that works. Notice the effect of the color scheme.

```{r}



library(gplots)
library(viridis)
heatmap.2(measles)
heatmap.2(measles, margins = c(3, 9))
heatmap.2(measles,trace = "none", margins = c(3, 9))
heatmap.2(measles, trace = "none", margins = c(3, 9), dendrogram = "row", Colv = NULL)
# heatmap.2(sqrt(measles), Colv = NULL, trace = "none", margins = c(3, 9))
library(viridis)
heatmap.2(measles, Colv = NULL, trace = "none", col = viridis(200), margins = c(3, 9))
heatmap.2(sqrt(measles), Colv = NULL, trace = "none", col = viridis(200), margins = c(3, 9))

# hist(measles, col = )

library(ggplot2)
library(viridis)
hp <- qplot(x = as.numeric(measles), fill = ..count.., geom = "histogram")
hp + scale_fill_viridis(direction = -1) + xlab("Measles incedence") + theme_bw()



```


Using scaling doesn't make much sense in this data.

```{r}

heatmap.2(measles, dendrogram = "row", Colv = NULL, trace = "none", margins = c(3, 9),
          col = viridis(200), 
          scale = "row"
          )
heatmap.2(measles, dendrogram = "row", Colv = NULL, trace = "none", margins = c(3, 9),
          col = viridis(200), 
          scale = "column"
          )



```


Clustering
-------------

Decisions for clustering (highlight_branches helps identify the topology of the dendrogram, it colors each branch based on its height):

```{r}
DATA <- measles
d <- dist(sqrt(DATA))

library(dendextend)
dend_expend(d)[[3]]
# average seems to make the most sense

dend_row <- d %>% hclust(method = "average") %>% as.dendrogram 
dend_row %>% highlight_branches %>% plot
```


What would be the optimal number of clusters? (it seems that 3 would be good)

```{r}

library(dendextend)

dend_k <- find_k(dend_row)
plot(dend_k)

```


Let's plot it:

```{r}

Rowv <- dend_row %>% color_branches(k = 3)
heatmap.2(sqrt(measles), Colv = NULL, Rowv = Rowv, trace = "none", col = viridis(200), margins = c(3, 9))

Rowv <- dend_row %>% color_branches(k = 3) %>% seriate_dendrogram(x=d)
heatmap.2(sqrt(measles), Colv = NULL, Rowv = Rowv, trace = "none", col = viridis(200), margins = c(3, 9))

```



heatmaply
-------------

All of this work (other then finding that the hclust method should be average) can simply be done using the following code to produce an interactive heatmap using heatmaply:


```{r, fig.width=12, fig.height=8}

library(heatmaply)
heatmaply(sqrt(measles), Colv = NULL, hclust_method = "average", 
          fontsize_row = 8,fontsize_col = 6,
          k_row = NA, margins = c(60,170, 70,40),
          xlab = "Year", ylab = "State", main = "Project Tycho's heatmap visualization \nof the measles vaccine"
        ) 


```

We can also use plot_method = "plotly", row_dend_left = TRUE
to make the plot more similar to heatmap.2 using the plotly engine (this is a bit more experimental. The default,  plot_method = "ggplot", is often safer to use - but it does not yet support row_dend_left = TRUE properly).

```{r, fig.width=12, fig.height=8}

heatmaply(sqrt(measles), Colv = NULL, hclust_method = "average", 
          fontsize_row = 8,fontsize_col = 6,
          k_row = NA, margins = c(60,50, 70,90),
          xlab = "Year", ylab = "State", main = "Project Tycho's heatmap visualization \nof the measles vaccine",
          plot_method = "plotly", row_dend_left = TRUE
        ) 

```


```{r, cache=FALSE}
# save.image("example.Rdata")
```




sessionInfo
===============


```{r}
sessionInfo()
```

